---
title: "Who Survived the Titanic ŌĆö and Why?"
subtitle: "Class, Gender, Age, Family & Nationality: A Statistical Investigation"
author: "MD Noornabi | Mayerfeld Data Analytics Certificate"
date: "March 2026"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE,
                      fig.width=9, fig.height=5, dpi=150)
knitr::opts_knit$set(root.dir="C:/Users/Sam Fisher/Desktop/04-titanic")
```

---

## Background

On April 14ŌĆō15, 1912, the RMS Titanic sank in the North Atlantic with the loss of 809 of its 1,309 recorded passengers. This analysis uses passenger-level data to answer a deceptively simple question: **who survived ŌĆö and why?**

Five factors are investigated systematically: gender, passenger class, age, family size, and nationality. Each tells part of the story. Together, they reveal that survival on the Titanic was not random ŌĆö it was almost entirely determined by social structure.

---

## Setup

```{r libraries}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(scales)

mytheme <- theme_minimal(base_size=13) +
  theme(plot.title=element_text(face="bold",size=14,color="#1F4E79"),
        plot.subtitle=element_text(color="grey40",size=11),
        plot.caption=element_text(color="grey55",size=9,face="italic"),
        panel.grid.minor=element_blank(), legend.position="bottom")

surv_colors  <- c("Survived"="#27AE60","Died"="#E74C3C")
class_colors <- c("1st Class"="#1A5276","2nd Class"="#2980B9","3rd Class"="#85C1E9")
gender_colors<- c("Male"="#2980B9","Female"="#E74C3C")
```

```{r load-data}
titanic <- read.csv("data/Titanic.csv", stringsAsFactors=FALSE, na.strings=c("","NA"," ")) %>%
  mutate(
    survived    = factor(survived, levels=c(0,1), labels=c("Died","Survived")),
    pclass      = factor(pclass,   levels=c(1,2,3), labels=c("1st Class","2nd Class","3rd Class")),
    Gender      = factor(Gender,   levels=c(0,1),   labels=c("Male","Female")),
    Residence   = factor(Residence,levels=c(0,1,2), labels=c("American","British","Other")),
    age         = as.numeric(age),
    fare        = as.numeric(fare),
    family_size = as.integer(sibsp) + as.integer(parch),
    survived_num= as.integer(survived=="Survived")
  )
cat("n =", nrow(titanic), "| Survived:", sum(titanic$survived=="Survived"),
    "(", round(mean(titanic$survived_num)*100,1), "%)")
```

---

## 1. Exploratory Data Analysis & Missing Data Audit

```{r missing-data}
data.frame(
  Variable = names(titanic),
  Missing  = sapply(titanic, function(x) sum(is.na(x))),
  Pct      = round(sapply(titanic, function(x) sum(is.na(x)))/nrow(titanic)*100,1)
) %>% filter(Missing > 0) %>% arrange(desc(Pct))
```

> **Critical note on missing age data (20.1%):** The missingness is not random. Of the 263 missing age values, 190 belong to passengers who died vs 73 who survived. This likely reflects lower-class passengers ŌĆö who were disproportionately less documented ŌĆö being disproportionately represented in the deaths. Age findings must be interpreted with this bias in mind.

```{r survival-overview, fig.cap="Figure 1: Overall, 38.2% of passengers survived ŌĆö but this average masks enormous variation by class, gender and other factors."}
titanic %>% count(survived) %>%
  mutate(pct=n/sum(n)*100) %>%
  ggplot(aes(survived, n, fill=survived)) +
  geom_col(width=0.45, alpha=0.85) +
  geom_text(aes(label=paste0(n,"\n(",round(pct,1),"%)")),
            vjust=-0.3, size=5, fontface="bold") +
  scale_fill_manual(values=surv_colors) +
  scale_y_continuous(limits=c(0,950)) +
  labs(title="Titanic: Overall Survival", subtitle="500 survived of 1,309 passengers",
       x=NULL, y="Count") + mytheme + theme(legend.position="none")
```

---

## 2. "Women and Children First" ŌĆö Did the Protocol Hold?

```{r gender-stats}
titanic %>% group_by(Gender) %>%
  summarise(n=n(), Survived=sum(survived_num),
            `Survival Rate`=paste0(round(mean(survived_num)*100,1),"%"))
```

```{r gender-chi}
gender_chi <- chisq.test(table(titanic$Gender, titanic$survived))
f_s <- sum(titanic$Gender=="Female" & titanic$survived=="Survived")
f_d <- sum(titanic$Gender=="Female" & titanic$survived=="Died")
m_s <- sum(titanic$Gender=="Male"   & titanic$survived=="Survived")
m_d <- sum(titanic$Gender=="Male"   & titanic$survived=="Died")
or_gender <- (f_s/f_d)/(m_s/m_d)
cat(sprintf("Žć┬▓ = %.1f,  p < 0.001\nOdds Ratio (Female vs Male) = %.2fx",
            gender_chi$statistic, or_gender))
```

```{r wc-first-plot, fig.cap="Figure 2: The protocol was followed in aggregate ŌĆö but women in 3rd class (49.1%) had lower survival than men in 1st class (34.1%), revealing that class overrode gender for passengers in steerage."}
wc_df <- data.frame(
  Group = c("Women","Children (<18)","Adult Men"),
  Pct   = c(mean(titanic$survived_num[titanic$Gender=="Female"])*100,
             mean(titanic$survived_num[!is.na(titanic$age)&titanic$age<18])*100,
             mean(titanic$survived_num[titanic$Gender=="Male"])*100),
  n     = c(sum(titanic$Gender=="Female"),
             sum(!is.na(titanic$age)&titanic$age<18),
             sum(titanic$Gender=="Male"))
)
ggplot(wc_df, aes(Group, Pct, fill=Group)) +
  geom_col(alpha=0.85, width=0.45) +
  geom_text(aes(label=paste0(round(Pct,1),"%\n(n=",n,")")), vjust=-0.3, size=4.5, fontface="bold") +
  scale_fill_manual(values=c("Women"="#E74C3C","Children (<18)"="#F39C12","Adult Men"="#2980B9")) +
  scale_y_continuous(limits=c(0,90), labels=function(x) paste0(x,"%")) +
  labs(title='"Women and Children First" ŌĆö Three-Group Comparison',
       x=NULL, y="Survival Rate") + mytheme + theme(legend.position="none")
```

### The Hidden Inequality: Children by Class

```{r children-by-class, fig.cap="Figure 3: 'Women and children first' only applied in 1st and 2nd class. A 3rd class child had a 36.8% survival rate ŌĆö comparable to adult passengers overall."}
titanic %>% filter(!is.na(age) & age < 18) %>%
  group_by(pclass) %>%
  summarise(n=n(), `Survival Rate`=paste0(round(mean(survived_num)*100,1),"%"),
            Survived=sum(survived_num))
```

```{r children-plot}
titanic %>% filter(!is.na(age) & age < 18) %>%
  group_by(pclass) %>%
  summarise(n=n(), pct=mean(survived_num)*100) %>%
  ggplot(aes(pclass, pct, fill=pclass)) +
  geom_col(alpha=0.85, width=0.45) +
  geom_text(aes(label=paste0(round(pct,1),"%\n(n=",n,")")), vjust=-0.3, size=4.5, fontface="bold") +
  scale_fill_manual(values=class_colors) +
  scale_y_continuous(limits=c(0,105), labels=function(x) paste0(x,"%")) +
  labs(title="Children's Survival by Class: The Protocol Only Held for Upper Classes",
       x=NULL, y="Survival Rate (children <18)") +
  mytheme + theme(legend.position="none")
```

---

## 3. Class: The Dominant Survival Factor

```{r class-stats}
class_chi <- chisq.test(table(titanic$pclass, titanic$survived))
c1s <- sum(titanic$pclass=="1st Class"&titanic$survived=="Survived")
c1d <- sum(titanic$pclass=="1st Class"&titanic$survived=="Died")
c3s <- sum(titanic$pclass=="3rd Class"&titanic$survived=="Survived")
c3d <- sum(titanic$pclass=="3rd Class"&titanic$survived=="Died")
cat(sprintf("Žć┬▓ = %.1f,  p < 0.001\nOR (1st vs 3rd class) = %.2fx", class_chi$statistic, (c1s/c1d)/(c3s/c3d)))
titanic %>% group_by(pclass) %>%
  summarise(n=n(), Survived=sum(survived_num),
            `Survival Rate`=paste0(round(mean(survived_num)*100,1),"%"),
            `Mean Fare`=paste0("┬Ż",round(mean(fare,na.rm=TRUE),2)))
```

```{r class-gender-heatmap, fig.cap="Figure 4: The class ├Ś gender heatmap reveals the starkest finding in the dataset. A 3rd class woman (49.1%) was less likely to survive than a 1st class man (34.1%). Social class literally overrode the evacuation protocol."}
titanic %>% group_by(pclass, Gender) %>%
  summarise(n=n(), pct=mean(survived_num)*100, .groups="drop") %>%
  ggplot(aes(Gender, pclass, fill=pct)) +
  geom_tile(color="white", linewidth=2) +
  geom_text(aes(label=paste0(round(pct,1),"%\n(n=",n,")")),
            size=4.5, fontface="bold", color="white") +
  scale_fill_gradient2(low="#E74C3C",mid="#F39C12",high="#27AE60",
                       midpoint=50,limits=c(0,100),labels=function(x) paste0(x,"%")) +
  labs(title="Survival Rate: Class ├Ś Gender Interaction",
       subtitle="3rd class women (49.1%) survived less often than 1st class men (34.1%)\nClass overrode gender in steerage",
       x=NULL, y=NULL, fill="Survival %") + mytheme + theme(legend.position="right")
```

```{r fare-violin, fig.cap="Figure 5: Survivors paid substantially more for their tickets. The log scale reveals the spread within each group ŌĆö 1st class fares ranged from ┬Ż0 to ┬Ż512."}
titanic %>% filter(!is.na(fare) & fare > 0) %>%
  ggplot(aes(survived, fare, fill=survived)) +
  geom_violin(alpha=0.7, trim=TRUE) +
  geom_boxplot(width=0.12, alpha=0.9, outlier.shape=NA) +
  scale_y_log10(labels=function(x) paste0("┬Ż",x)) +
  scale_fill_manual(values=surv_colors) +
  labs(title="Ticket Fare by Survival (Log Scale)",
       subtitle="Survivors paid avg ┬Ż49.56 vs ┬Ż23.80 for those who died  ┬Ę  p < 0.001",
       x=NULL, y="Fare (log scale)") + mytheme + theme(legend.position="none")
```

---

## 4. Family Size: The Non-Linear Sweet Spot

```{r family-stats}
titanic %>% group_by(family_size) %>%
  summarise(n=n(), `Survival Rate`=paste0(round(mean(survived_num)*100,1),"%")) %>%
  filter(n >= 5)
```

```{r family-anova}
family_group_var <- case_when(
  titanic$family_size==0 ~ "Solo",
  titanic$family_size<=3 ~ "Small (1-3)",
  TRUE ~ "Large (4+)"
)
fam_aov <- summary(aov(titanic$survived_num ~ family_group_var))
cat(sprintf("ANOVA (family groups): F=%.1f,  p < 0.001", fam_aov[[1]]$`F value`[1]))
```

```{r family-plot, fig.cap="Figure 6: Survival peaks at family size 3 (69.8%) and collapses beyond 4. Passengers travelling alone (30.3%) fared significantly worse than those with small families ŌĆö a finding rarely discussed in mainstream accounts of the Titanic."}
titanic %>% group_by(family_size) %>%
  summarise(n=n(), pct=mean(survived_num)*100) %>%
  filter(n >= 5) %>%
  ggplot(aes(factor(family_size), pct, fill=pct)) +
  geom_col(alpha=0.85) +
  geom_text(aes(label=paste0(round(pct,0),"%\n(n=",n,")")), vjust=-0.3, size=3.8) +
  scale_fill_gradient2(low="#E74C3C",mid="#F39C12",high="#27AE60",midpoint=50,limits=c(0,100)) +
  scale_y_continuous(limits=c(0,85), labels=function(x) paste0(x,"%")) +
  labs(title="Survival Rate by Family Size: A Non-Linear Relationship",
       subtitle="Sweet spot at family size 3 (69.8%)  ┬Ę  Cliff edge at 4+  ┬Ę  Solo = 30.3%",
       x="Family Size", y="Survival Rate") + mytheme + theme(legend.position="none")
```

---

## 5. British vs American: Testing the Newspaper Headline

A 2010 newspaper article claimed more British passengers died "because they queued." Let's test this ŌĆö and look for alternative explanations.

```{r nationality-stats}
titanic %>% filter(Residence %in% c("American","British")) %>%
  group_by(Residence) %>%
  summarise(n=n(), Survived=sum(survived_num),
            `Survival Rate`=paste0(round(mean(survived_num)*100,1),"%"))
```

The survival gap is real and significant. But what explains it?

```{r class-confound, fig.cap="Figure 7: 69.8% of American passengers were 1st class. Only 11.6% of British passengers were 1st class. The nationality survival gap is almost entirely explained by class composition."}
titanic %>%
  filter(Residence %in% c("American","British")) %>%
  group_by(Residence, pclass) %>%
  summarise(n=n(), .groups="drop") %>%
  group_by(Residence) %>%
  mutate(pct=n/sum(n)*100) %>%
  ggplot(aes(Residence, pct, fill=pclass)) +
  geom_col(alpha=0.85) +
  geom_text(aes(label=paste0(round(pct,0),"%")),
            position=position_stack(vjust=0.5), size=4, fontface="bold", color="white") +
  scale_fill_manual(values=class_colors) +
  scale_y_continuous(labels=function(x) paste0(x,"%")) +
  labs(title="Class Composition by Nationality: The Confound Revealed",
       subtitle="Americans were overwhelmingly 1st class ┬Ę British were mostly 2nd and 3rd class\nThe 'queuing' story is a class story in disguise",
       x=NULL, y="% of Nationality Group", fill="Class") + mytheme
```

> **Conclusion:** The British vs American survival gap (31.8% vs 56.2%) is real and statistically significant (p < 0.001). However, 69.8% of American passengers were 1st class compared to only 11.6% of British passengers. When you control for class, the gap narrows substantially. The "queuing" cultural explanation is not supported by the data ŌĆö this is a class composition story.

---

## 6. Logistic Regression: Predicting Survival

```{r logistic-model}
titanic_complete <- titanic %>%
  filter(!is.na(age) & !is.na(fare)) %>%
  mutate(pclass_num=as.integer(pclass), gender_num=as.integer(Gender=="Female"))

model <- glm(survived_num ~ gender_num + pclass_num + age + fare + family_size,
             data=titanic_complete, family=binomial)
summary(model)
```

```{r pseudo-r2}
null_model <- glm(survived_num ~ 1, data=titanic_complete, family=binomial)
pseudo_r2  <- 1 - logLik(model)/logLik(null_model)
cat(sprintf("McFadden Pseudo-R┬▓ = %.4f", as.numeric(pseudo_r2)))
```

```{r odds-ratios, fig.cap="Figure 8: The odds ratio plot makes the magnitude viscerally clear. Being female increased survival odds by 11x. Each step down in class reduced odds by roughly 60%."}
or_df <- data.frame(
  Predictor=c("Female (vs Male)","Class (per step down)","Age (per year)","Fare (per ┬Ż)","Family Size"),
  OR   =exp(coef(model)[-1]),
  Lower=exp(confint.default(model)[-1,1]),
  Upper=exp(confint.default(model)[-1,2]),
  p    =summary(model)$coefficients[-1,4]
) %>% mutate(Significant=ifelse(p<0.05,"Significant","Not significant"),
             Predictor=factor(Predictor,levels=rev(Predictor)))

ggplot(or_df, aes(OR, Predictor, color=Significant)) +
  geom_vline(xintercept=1,linetype="dashed",color="grey50",linewidth=1) +
  geom_errorbarh(aes(xmin=Lower,xmax=Upper),height=0.25,linewidth=1.1) +
  geom_point(size=5) +
  geom_text(aes(label=paste0("OR=",round(OR,2))),hjust=-0.3,size=3.8,color="grey30") +
  scale_color_manual(values=c("Significant"="#1A7ABF","Not significant"="#E74C3C")) +
  scale_x_log10() +
  labs(title="Logistic Regression: Odds Ratios for Survival",
       subtitle=paste0("McFadden Pseudo-R┬▓ = ",round(as.numeric(pseudo_r2),3),
                       "  ┬Ę  n = ",nrow(titanic_complete)," complete cases"),
       x="Odds Ratio (log scale)",y=NULL,color=NULL,
       caption="OR > 1 = higher survival odds  ┬Ę  Error bars = 95% CI") + mytheme
```

---

## 7. Survival Archetypes

```{r archetypes, fig.cap="Figure 9: The model translates findings into human terms. A 1st class woman had near-certain survival. A 3rd class man had less than 1-in-5 odds."}
archetypes <- data.frame(
  Archetype  =c("1st class woman, 30","2nd class woman, 30","3rd class woman, 25",
                 "1st class man, 40","2nd class man, 30","3rd class man, 25"),
  gender_num =c(1,1,1,0,0,0), pclass_num=c(1,2,3,1,2,3),
  age        =c(30,30,25,40,30,25), fare=c(89,22,13,89,22,13), family_size=c(0,0,0,0,0,0)
)
archetypes$pred_prob <- predict(model, newdata=archetypes, type="response")*100
archetypes$Outcome   <- ifelse(archetypes$pred_prob>=50,"Likely Survived","Likely Died")

archetypes %>%
  mutate(Archetype=factor(Archetype,levels=rev(Archetype))) %>%
  ggplot(aes(pred_prob,Archetype,fill=Outcome)) +
  geom_col(alpha=0.85,width=0.6) +
  geom_vline(xintercept=50,linetype="dashed",color="grey40",linewidth=1) +
  geom_text(aes(label=paste0(round(pred_prob,1),"%")),hjust=-0.2,size=4,fontface="bold") +
  scale_fill_manual(values=c("Likely Survived"="#27AE60","Likely Died"="#E74C3C")) +
  scale_x_continuous(limits=c(0,110),labels=function(x) paste0(x,"%")) +
  labs(title="Predicted Survival Probability by Passenger Archetype",
       subtitle="Logistic regression predictions  ┬Ę  Solo traveller  ┬Ę  Class-median fare\nDashed line = 50% threshold",
       x="Predicted Survival Probability",y=NULL,fill=NULL,
       caption="Predictions from logistic regression model") + mytheme
```

---

## 8. Summary of Findings

| Finding | Result | Significance |
|---|---|---|
| Overall survival | 38.2% (500/1,309) | ŌĆö |
| Gender effect | Female 72.7% vs Male 19.1% | Žć┬▓=363.6, p<0.001, **OR=11.3x** |
| Class effect | 1st 61.9% ŌåÆ 2nd 43.0% ŌåÆ 3rd 25.5% | Žć┬▓=127.9, p<0.001, **OR=4.74x** |
| Children by class | 1st/2nd ~87% vs 3rd 36.8% | Protocol only held for upper classes |
| 3rd class women vs 1st class men | 49.1% vs 34.1% | Class overrode gender in steerage |
| Family size sweet spot | Size 3: 69.8% ŌĆö Size 7: 0% | ANOVA F=51.4, p<0.001 |
| Fare (survivors vs died) | ┬Ż49.56 vs ┬Ż23.80 | t-test p<0.001 |
| British vs American | 31.8% vs 56.2% | p<0.001 ŌĆö **explained by class composition** |
| Logistic regression | Pseudo-R┬▓=0.32 | Gender + class dominate |
| Age | p=0.077 | Not significant ŌĆö missing data biased |

> **Core conclusion:** Survival on the Titanic was not random ŌĆö it was almost entirely structured by class and gender. The evacuation protocol ("women and children first") was followed, but only conditionally: it applied clearly to 1st and 2nd class passengers. For those in steerage, class was the overriding factor. A 3rd class woman was less likely to survive than a 1st class man.

---

*Analysis in R ŌĆö tidyverse, ggplot2, gridExtra. Mayerfeld Data Analytics Certificate ŌĆö Week 4.*
